Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-16 manuscript/climate_sensitivity_draft.Rmd
Consolidated comments up to end of 3rd paragraph of intro
#65

\newpage
### Introduction
**Nobby says this intro is based too much on Bennet 2015. I'm not sure if he means we should remove some of the citations or rephrase some sentences.**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014] **Neil adds in a number of refs here, unsure if we should just include all of them**, sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@allen_global_2010; **REFS**] **Alan suggests not citing Allen2010 here and instead gives a list of other sources. Erika also suggests 2 papers**. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic response. That study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper. Bill says this sentence is confusing but I believe it is fine**. First, they found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. **For this previous sentence, Alan questions whether this assumption can be made, and Valentine says it's odd to come back to Bennett as a ref here.** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **Valentine is confused if you mean to say "lack of" these things. Alan points out we should define these terms here since is the first time they're used. Unsure what your thoughts are on that**. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.

A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response. Xylem architecture plays a role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but does not differentiate species beyond broad classes. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018). Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the osmotic potential at turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. The ability of $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

>>>>

\newpage
### Introduction
**Nobby says this intro is based too much on Bennet 2015. I'm not sure if he means we should remove some of the citations or rephrase some sentences.**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, they quantified tree size based on DBH, which has no direct mechanistic response. Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper; I'm unsure how you want to consolidate that with this sentence.**. First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. **For this previous sentence, Alan questions whether this assumption can be made, and Valentine says it's odd to come back to Bennett as a ref here.** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **Valentine is confused if you mean to say "lack of" these things. Alan points out we should define these terms here since is the first time they're used. Unsure what your thoughts are on that**. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.

A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response [@elliott_forest_2015; @martinbenito_convergence_2015]. Separating species by ring porosity provides a first step toward identifying sensitivity, with diffuse-porous species tending to be more sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but additional insight similar to @roskilly_conflicting_2019 is needed to better predict sensitivity at species or individual levels. Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017]. However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018). Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019]. These include traits such as the fraction of conductivity lost to embolism, is another relevant metric (percent loss conductivity) [**REF??**], leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]), and the leaf water potential at turgor loss point ($\pi_{tlp}$), i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
